Understanding Recovery from Malnutrition in Early Life:\\ A Multi-Omic Analysis of the Infant Gut Microbiome, Plasma Lipidome, and Neurodevelopmental Outcomes

### 2025-08-03 writing/manuscript/main.tex
mtk and manuscript
mtk and rda analysis

\section*{Results}
\subsection*{Study population characteristics}
As a city with the second highest density of population and in a country with childhood malnutrition rate is one of the highest globally, the Mirpur region in Dhaka, Bangladesh was chosen to assess the impact of early-life malnutrition \cite{ahmed2012nutrition}.
Refeeding resulted in recovery in approximately half the children with MAM, with no significant difference in rate between the two refeed methods (43.0\% in the Local RUSF group vs 41.2\% in the ERUSF group , MWU. p = 0.9460).
However, a key distinction emerged in time to recovery, which was significantly shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{../../figures/table1.pdf}
	\caption[Study population characteristics and recovery rates]{
		Study population characteristics and recovery rates.
	}
\label{Figure1}
\end{figure}

\subsection*{Impact of the Enhanced Refeed regimen on Neurodevelopmental Outcomes}
Using mixed-effects models controlling for sex and delivery mode, the enhanced refeed intervention showed no statistically significant main effects or interactions with timepoint across cognitive, communication, and motor Bayley scores (Table S2).
Enhanced refeeding did not significantly improve most cognitive/behavioral outcomes, though some marginal trends in fine motor scores were observed.
While children improved in their cognitive abilities over time, this improvement did not differ significantly by intervention group.

\begin{figure}[H]
\centering
\includegraphics[scale=1.2]{../../figures/anthrochange_simple.pdf}
	\caption[Measures of cognition were not significantly impacted by refeeding]{
		Measures of cognition were not significantly impacted by refeeding.
		a) Lineplot of how children change in WLZ/WHZ during refeeding.
		b) Boxplot of days to recovery compared between the feeding groups.
		c) Cognitive factors that were associated with refeeding (Left and right heatmaps of Bayley score and Wolkes respectively).
		d) Lineplot of change of Expressive Communication Bayley Score stratified by Well-nourished, ERUSF, or RUSF.
		e) Boxplot of change of vocalisation Wolkes score stratified by refeed type.
	}
\label{Figure2}
\end{figure}

\subsection*{Baseline factors that influenced Bayley scores}
With the primary outcome investigated, a post-hoc, discovery approach into the factors that influenced cognition was performed.
Random Forest regression analysis to predict Bayley scores at the baseline and two-year timepoint from the baseline datasets revealed that, although Bayley Scores were largely unexplainable from the baseline datasets, Polygenetic Risk Score (PRS) factors were predictive of expressive communication scores at the baseline timepoint.
This predictive power was lost at the year-two timepoint.
The PRS score that was most differentially associated with expressive communication score was the Intelligence PRS.
\begin{figure}[H]
	\centering
	\includegraphics[scale=1.2]{../../figures/predictbayley.pdf}
	\caption[Factors that were predictive of Bayley scores]{
		Factors that were predictive of Bayley scores.
	}
\label{Figure1}
\end{figure}

\subsection*{Baseline Predictors of Recovery}
Acknowleging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
Anthropometric and genetic indicators were stronger predictors than microbiome or EEG.
The most significant factors were anthropometric such as head circumference, indicating extent of malnutrition.
Other factors included the proportion of days without feeding failure, indicating compliance/adherance to the refeeding.
\begin{figure}[H]
\centering
\includegraphics[scale=0.8]{../../figures/ev.pdf}
	\caption[Factors that associate with recovery.]{
		Factors that associate with recovery.
		Explained variance as measured by PERMANOVA of beta diversities between samples within each dataset.
	}
\label{Figure3}
\end{figure}

\subsection*{Assessing the impacts with respect to well-nourished controls}
The classification of a given microbiome as 'healthy' has long been contested.
This is true also of a number of datasets within this cohort.
Thus, a method whereby the distance between the composition of the overall community of each dataset was compared with the respective well-nourished age-matched control for each timepoint.
For each dataset analysed through this techhniqe, the shrinking/widening of this distance with respect to the well-nourished controls between the two timepoints was used to indicate improvement/deterioration.
ERUTF drove a greater shift toward healthy microbiome, including depletion of S. wadsworthensis and L. fermentum.
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{../../figures/newrecoveryfeed3.pdf}
	\caption[Refeed methods improve/deteriorate bodily systems differently]{
		Refeed methods improve/deteriorate bodily systems differently.
		A, Radial plot of percentage change in distance from Well-nourished controls before and after refeeding (1yr and 2yr). Positive and negative percentages represent improvement and deterioration, respectively.
		B, Horizontal barplot of features that explain the change significantly (q < 0.2) between refeed methods after mixed modelling for each dataset.
		C, Heatmap to show clusters of dataset changes.
		D, Boxplot to show how Lactobacillus fermentum changes according to the refeed method.
	}
\label{Figure1}
\end{figure}

\subsection*{Network-Based Integration of Multimodal Changes}
With an understanding of the impacts that the enhanced refeed method had on developmental trajectories across the datasets recoreded, the next step was to understand how those changes might be connected together mechanisticly.
It was found that proline metabolism was correlated with EEG and cognitive outcomes, suggesting neuro-metabolic recovery pathways.
\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{../../figures/mixedmodelnetwork.pdf}
	\caption[Mechanism of E-RUTF differential impacts on brain structure]{
		Mechanism of E-RUTF differential impacts on brain structure.
	}
\label{Figure4}
\end{figure}

@@

\section*{Methods}
PSS method \cite{mozumder2022reliability}
\subsection*{Ethics}
The M4EFaD intervention was registered NCT05629624 on clinicaltrials.gov.
The study was approved by icddr,b Ethical Review Committee PR-21084 and the Bangladesh Directorate General of Drug Administration.
Ethical review for the analytical component was obtained from Auckland Health Research Ethics Committee approval AH23922 (metabolomics, metagenomics, machine learning).

@@

\bibliography{library}

@@

\section*{Supplementary material}
\begin{figure}[!htb]
\centering
\includegraphics[scale=0.8]{../../figures/newalldata.pdf}
	\caption[Datasets recoreded in the M4EFaD study]{
		Datasets recorded in the M4EFaD study.
	}
\label{Figure1}
\end{figure}

>>>>

\section*{Results}
\subsection*{Study population characteristics}
As a city with the second highest density of population and in a country with childhood malnutrition rate is one of the highest globally, the Mirpur region in Dhaka, Bangladesh was chosen to assess the impact of early-life malnutrition \cite{ahmed2012nutrition}.
Refeeding resulted in recovery in approximately half the children with MAM, with no significant difference in rate between the two refeed methods (43.0\% in the Local RUSF group vs 41.2\% in the ERUSF group , MWU. p = 0.9460).
However, a key distinction emerged in time to recovery, which was shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{figures/table1.pdf}
	\caption[Study population characteristics and recovery rates]{
		Study population characteristics and recovery rates.
	}
\label{Figure1}
\end{figure}

\subsection*{Impact of the Enhanced Refeed regimen on Neurodevelopmental Outcomes}
Using mixed-effects models controlling for sex and delivery mode, the enhanced refeed intervention showed no statistically significant main effects or interactions with timepoint across cognitive, communication, and motor Bayley scores (Table S2).
Enhanced refeeding did not significantly improve most cognitive/b shavioral outcomes, though some marginal trends in fine motor scores were observed.
While children improved in their cognitive abilities over time, this improvement did not differ significantly by intervention group.

\begin{figure}[H]
\centering
\includegraphics[scale=1.2]{figures/anthrochange_simple.pdf}
	\caption[Measures of cognition were not significantly impacted by refeeding]{
		Measures of cognition were not significantly impacted by refeeding.
		a) Lineplot of how children change in WLZ/WHZ during refeeding.
		b) Boxplot of days to recovery compared between the feeding groups.
		c) Cognitive factors that were associated with refeeding (Left and right heatmaps of Bayley score and Wolkes respectively).
		d) Lineplot of change of Expressive Communication Bayley Score stratified by Well-nourished, ERUSF, or RUSF.
		e) Boxplot of change of vocalisation Wolkes score stratified by refeed type.
	}
\label{Figure2}
\end{figure}

\subsection*{Baseline factors that influenced Bayley scores}
With the primary outcome investigated, a post-hoc, discovery approach into the factors that influenced cognition was performed.
Random Forest regression analysis to predict Bayley scores at the baseline and two-year timepoint from the baseline datasets revealed that, although Bayley Scores were largely unexplainable from the baseline datasets, Polygenetic Risk Score (PRS) factors were predictive of expressive communication scores at the baseline timepoint.
This predictive power was lost at the year-two timepoint.
The PRS score that was most differentially associated with expressive communication score was the Intelligence PRS.

\subsection*{Baseline Predictors of Recovery}
Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
Anthropometric and genetic indicators were stronger predictors than microbiome or EEG.
The most significant factors were anthropometric such as head circumference, indicating extent of malnutrition.
Other factors included the proportion of days without feeding failure, indicating compliance/adherance to the refeeding.
\begin{figure}[H]
\centering
\includegraphics[scale=0.8]{figures/ev.pdf}
	\caption[Factors that associate with recovery.]{
		Factors that associate with recovery.
		Explained variance as measured by PERMANOVA of beta diversities between samples within each dataset.
	}
\label{Figure3}
\end{figure}

\subsection*{Assessing the impacts with respect to well-nourished controls}
The classification of a given microbiome as 'healthy' has long been contested.
This is true also of a number of datasets within this cohort.
Thus, a method whereby the distance between the composition of the overall community of each dataset was compared with the respective well-nourished age-matched control for each timepoint.
For each dataset analysed through this techhniqe, the shrinking/widening of this distance with respect to the well-nourished controls between the two timepoints was used to indicate improvement/deterioration.
ERUTF drove a greater shift toward healthy microbiome, including depletion of S. wadsworthensis and L. fermentum.
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{figures/newrecoveryfeed3.pdf}
	\caption[Refeed methods improve/deteriorate bodily systems differently]{
		Refeed methods improve/deteriorate bodily systems differently.
		A, Radial plot of percentage change in distance from Well-nourished controls before and after refeeding (1yr and 2yr). Positive and negative percentages represent improvement and deterioration, respectively.
		B, Horizontal barplot of features that explain the change significantly (q < 0.2) between refeed methods after mixed modelling for each dataset.
		C, Heatmap to show clusters of dataset changes.
		D, Boxplot to show how Lactobacillus fermentum changes according to the refeed method.
	}
\label{Figure1}
\end{figure}

\subsection*{Network-Based Integration of Multimodal Changes}
With an understanding of the impacts that the enhanced refeed method had on developmental trajectories across the datasets recoreded, the next step was to understand how those changes might be connected together mechanisticly.
It was found that proline metabolism was correlated with EEG and cognitive outcomes, suggesting neuro-metabolic recovery pathways.
\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{figures/mixedmodelnetwork.pdf}
	\caption[Mechanism of E-RUTF differential impacts on brain structure]{
		Mechanism of E-RUTF differential impacts on brain structure.
	}
\label{Figure4}
\end{figure}

@@

\section*{Methods}
PSS method \cite{mozumder2022reliability}

\subsection*{Ethics}
The M4EFaD intervention was registered NCT05629624 on clinicaltrials.gov.
The study was approved by icddr,b Ethical Review Committee PR-21084 and the Bangladesh Directorate General of Drug Administration.
Ethical review for the analytical component was obtained from Auckland Health Research Ethics Committee approval AH23922 (metabolomics, metagenomics, machine learning).

@@

\bibliography{writing/manuscript/library}

@@

\section*{Supplementary material}
\begin{figure}[!htb]
\centering
\includegraphics[scale=0.8]{figures/newalldata.pdf}
	\caption[Datasets recoreded in the M4EFaD study]{
		Datasets recorded in the M4EFaD study.
	}
\label{Figure1}
\end{figure}
